---
title: "The Q-Value for Multiple Comparison Correction"
author: "Nick Strayer"
date: "April 24, 2015"
output: html_document
---

We will be using the algorithm for calculating the q-value for a given p-value given an array of p-values from Storey and Tibshiran, 2003.

```{r echo = FALSE}
library(fdrtool)
library(ggplot2)
library(grid)
library(gridExtra)
library(plyr)
library(pspline)
```


--- 

## First we load in the data from the hedonometer paper: 

```{r}
setwd("/Users/Nick/spring15/independentStudy")#have to set this by chunk of code
obesityRaw = read.table("data/obesity_censored.txt", header = TRUE, fill = TRUE)
obesity = na.omit(obesityRaw) #There are some NAs in the data.
head(obesity)
```

Now we will set up an object that we will use throughout the process: 

```{r, fig.align='center'}
tests = data.frame(obesity$WordHappiness, obesity$pvalue)
names(tests) = c("word", "pval")
head(tests)
```
```{r echo = FALSE, fig.align='center'}

ggplot(tests, aes(x = pval)) + 
  geom_histogram(aes(y=..count../sum(..count..)), color = "black", fill = "#7fc97f") + 
  labs(title = "Histogram of p-values for obesity words (normalized)",
      x = "P-Value", y = "% of words at p-val")
```

## Now the algorithm: 

### 1) Order the p-values. 
__Note:__ The data are already in order, but I will include the step anyways.
```{r}
tests = arrange(tests, pval)
head(tests)
```

### 2) Look at the potential $\hat{\pi_0}(\lambda)$ values: 

$\hat{\pi_0}(\lambda)=\frac{\#\{p_i>\lambda\}}{m(1-\lambda)}$
```{r, fig.align='center'}
#Generate lambdas
m = length(tests$pval)
lambdas = seq(0.01,0.95,0.01)
piHat   = sapply(lambdas, function(l) sum(tests$pval > l)/m*(1 - l) )

piHat_df = data.frame(lambdas, piHat)

ggplot(piHat_df,aes(x=lambdas, y=piHat)) + geom_point() 
```

### 3) Fit a spline to the $\hat{\pi_0}$ values
__Note__: Not super sure why the spline is so weird. Investigating further methods of modeling this trend could be useful. 


```{r, fig.align='center'}
spline = smooth.Pspline(lambdas,piHat, norder = 3, df = 3, method = 2)

piHat_df$spline = spline$ysmth

ggplot(piHat_df,aes(x=lambdas)) + geom_point(aes(y = piHat)) + geom_line(aes(y = spline))
```

### 4) Set $\hat{\pi_0}$ to the spline at 1. 

```{r}
pi0 = predict(spline, 1)
print(pi0)
```

### 5) Calculate $\hat{q_{m}}$
$$\hat{q_{m}} = \hat{\pi_o} \cdot p_m$$
```{r}
#Start the array of values by putting in the last one
qvals = pi0 * tail(tests$pval, 1)
```

### 6) Find all of the rest of the q vals: 

$$min((\hat{\pi_o} \cdot m \cdot p_i)/i, q_{i + 1})$$

```{r}

for(i in (m-1):1 ){
  
  latest = pi0 * m * tests$pval[i]/ i
  last   = qvals[1] #because we are filling qvals reverse, 1 will always be the last computed val
  
  q = min(latest, last)
  qvals = c(q, qvals)
}

tests$qval = qvals

head(tests)
```


---
### Point of note: 

* I think it would be interesting to look more into the distribution of $\hat{pi}$ and also the type of curve fitted to it. 

